---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

1. Load the data (i.e. `read.csv()`)

2. Process/transform the data (if necessary) into a format suitable for your analysis

```{r}
if(!file.exists("activity.zip"))
  download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", "activity.zip")

if (!file.exists("activity.csv"))
  unzip("activity.zip")

data <- read.csv("activity.csv")
```


## What is mean total number of steps taken per day?

1. Make a histogram of the total number of steps taken each day

2. Calculate and report the **mean** and **median** total number of steps taken per day

```{r} 
stepsByDay <- aggregate(data["steps"], by=data["date"], sum)

meanSteps <- format(mean(stepsByDay$steps, na.rm="TRUE"), scientific = FALSE)

medianSteps <- median(stepsByDay$steps, na.rm=TRUE)

hist(stepsByDay$steps,
     xlab = "Total number of steps taken each day",
     ylab = "Count",
     main = "Histogram - total number of steps taken each",
     col = 4)
```

The mean total number of steps taken each day is `r meanSteps`, and the median number of steps taken each day is `r medianSteps`.


## What is the average daily activity pattern?
```{r}
library(magrittr)
library(dplyr)
## Get the avg numbers of steps taken for each 5 min interval

stepsByIntveral <- data %>% group_by(interval) %>% summarize(avgSteps = mean(steps, na.rm = TRUE))

## plot the time series of the avg number of steps taken per 5 min interval
plot(stepsByIntveral$interval, stepsByIntveral$avgSteps,
     xlab = "5 minute Interval",
     ylab = "Avg number of steps",
     main = "Time Series - Avg number of steps for each 5 min interval",
     col = 4)

```

1. Make a time series plot (i.e. `type = "l"`) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?


## Imputing missing values


## Are there differences in activity patterns between weekdays and weekends?
